---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd -->

# metANN

`metANN` is an R package for training feed-forward artificial neural networks using metaheuristic and gradient-based optimization algorithms. It also provides a general-purpose continuous optimization interface for benchmark functions and other numerical optimization problems.

The package supports:

- general-purpose continuous optimization,
- feed-forward multilayer perceptron training,
- regression,
- binary classification,
- multi-class classification,
- metaheuristic optimization,
- gradient-based optimization.

## Main Features

`metANN` currently supports the following optimizers:

```{r available-optimizers, eval=FALSE}
available_optimizers()
```

Currently available optimizers are:

```r
"pso"  "de"   "ga"   "abc"  "gwo"  "woa"  "tlbo" "sboa" "sgd"  "adam"
```

Metaheuristic optimizers:

```{r available-metaheuristics, eval=FALSE}
available_metaheuristics()
```

Gradient-based optimizers:

```{r available-gradient-optimizers, eval=FALSE}
available_gradient_optimizers()
```

## Installation

You can install the development version of `metANN` from GitHub using:

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("YOUR_GITHUB_USERNAME/metANN")
```

After installation, load the package:

```{r load-package, eval=FALSE}
library(metANN)
```

Replace `YOUR_GITHUB_USERNAME` with your GitHub username after uploading the package to GitHub.

## General-Purpose Optimization

The `met_optimize()` function can be used for continuous optimization problems.

Metaheuristic optimizers require only an objective function.

Gradient-based optimizers require both an objective function and its gradient.

## Example 1: Sphere Function with SBOA

```{r sphere-sboa, eval=FALSE}
sphere <- function(x) {
  sum(x^2)
}

result_sboa <- met_optimize(
  fn = sphere,
  optimizer = optimizer_sboa(
    pop_size = 30,
    max_iter = 100
  ),
  lower = rep(-10, 10),
  upper = rep(10, 10),
  seed = 123,
  verbose = FALSE
)

result_sboa
summary(result_sboa)
coef(result_sboa)

plot(result_sboa)
plot(result_sboa, log = TRUE)
```

## Example 2: Rastrigin Function with SBOA

```{r rastrigin-sboa, eval=FALSE}
rastrigin <- function(x) {
  10 * length(x) + sum(x^2 - 10 * cos(2 * pi * x))
}

result_rastrigin <- met_optimize(
  fn = rastrigin,
  optimizer = optimizer_sboa(
    pop_size = 40,
    max_iter = 150
  ),
  lower = rep(-5.12, 5),
  upper = rep(5.12, 5),
  seed = 123,
  verbose = FALSE
)

result_rastrigin
summary(result_rastrigin)

plot(result_rastrigin)
plot(result_rastrigin, log = TRUE)
```

## Gradient-Based General Optimization

Gradient-based optimizers such as SGD and Adam require a gradient function supplied via the `gr` argument.

## Example 3: Sphere Function with Adam

```{r sphere-adam, eval=FALSE}
sphere <- function(x) {
  sum(x^2)
}

grad_sphere <- function(x) {
  2 * x
}

result_adam <- met_optimize(
  fn = sphere,
  gr = grad_sphere,
  optimizer = optimizer_adam(
    learning_rate = 0.1,
    epochs = 100
  ),
  lower = rep(-5, 5),
  upper = rep(5, 5),
  initial = rep(4, 5),
  seed = 123,
  verbose = FALSE
)

result_adam
summary(result_adam)
coef(result_adam)

plot(result_adam)
plot(result_adam, log = TRUE)
```

## Example 4: Rastrigin Function with Adam

```{r rastrigin-adam, eval=FALSE}
rastrigin <- function(x) {
  10 * length(x) + sum(x^2 - 10 * cos(2 * pi * x))
}

grad_rastrigin <- function(x) {
  2 * x + 20 * pi * sin(2 * pi * x)
}

result_adam_rastrigin <- met_optimize(
  fn = rastrigin,
  gr = grad_rastrigin,
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 500
  ),
  lower = rep(-5.12, 5),
  upper = rep(5.12, 5),
  initial = rep(3, 5),
  seed = 123,
  verbose = FALSE
)

result_adam_rastrigin
summary(result_adam_rastrigin)

plot(result_adam_rastrigin)
```

Since the Rastrigin function is multimodal, gradient-based optimizers may converge to local minima depending on the initial point and learning rate.

## Feed-Forward Neural Network Training

The `met_mlp()` function is a convenient wrapper for training feed-forward multilayer perceptrons.

The task can be selected manually using:

```r
task = "regression"
task = "classification"
```

or detected automatically using:

```r
task = "auto"
```

When `task = "auto"`:

- numeric response variables are treated as regression,
- factor, character, or logical response variables are treated as classification.

## Example 5: MLP Regression with SBOA

```{r mlp-regression-sboa, eval=FALSE}
fit_reg_sboa <- met_mlp(
  formula = Petal.Width ~ Sepal.Length + Sepal.Width + Petal.Length,
  data = iris,
  hidden_layers = c(5),
  activation = "relu",
  optimizer = optimizer_sboa(
    pop_size = 20,
    max_iter = 30
  ),
  seed = 123,
  verbose = FALSE
)

fit_reg_sboa
summary(fit_reg_sboa)

evaluate(fit_reg_sboa, newdata = iris)

predict(fit_reg_sboa, iris[1:5, ], type = "response")

plot(fit_reg_sboa)
```

## Example 6: MLP Regression with Adam

```{r mlp-regression-adam, eval=FALSE}
fit_reg_adam <- met_mlp(
  formula = Petal.Width ~ Sepal.Length + Sepal.Width + Petal.Length,
  data = iris,
  hidden_layers = c(5),
  activation = "relu",
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 100,
    batch_size = 32
  ),
  seed = 123,
  verbose = FALSE
)

fit_reg_adam
summary(fit_reg_adam)

evaluate(fit_reg_adam, newdata = iris)

predict(fit_reg_adam, iris[1:5, ], type = "response")

plot(fit_reg_adam)
```

## Binary Classification

For binary classification, `metANN` uses a one-unit output layer with sigmoid activation by default.

If `loss = NULL`, the default loss is binary cross-entropy.

## Example 7: Binary Classification with SBOA

```{r binary-classification-sboa, eval=FALSE}
iris_bin <- iris

iris_bin$IsSetosa <- factor(
  ifelse(iris_bin$Species == "setosa", "setosa", "other")
)

fit_bin_sboa <- met_mlp(
  formula = IsSetosa ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = iris_bin,
  hidden_layers = c(5),
  activation = "relu",
  optimizer = optimizer_sboa(
    pop_size = 20,
    max_iter = 30
  ),
  seed = 123,
  verbose = FALSE
)

fit_bin_sboa
summary(fit_bin_sboa)

evaluate(fit_bin_sboa, newdata = iris_bin)

predict(fit_bin_sboa, iris_bin[1:5, ], type = "class")

predict(fit_bin_sboa, iris_bin[1:5, ], type = "prob")

plot(fit_bin_sboa)
```

## Example 8: Binary Classification with Adam

```{r binary-classification-adam, eval=FALSE}
iris_bin <- iris

iris_bin$IsSetosa <- factor(
  ifelse(iris_bin$Species == "setosa", "setosa", "other")
)

fit_bin_adam <- met_mlp(
  formula = IsSetosa ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = iris_bin,
  hidden_layers = c(5),
  activation = "relu",
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 100,
    batch_size = 32
  ),
  seed = 123,
  verbose = FALSE
)

fit_bin_adam
summary(fit_bin_adam)

evaluate(fit_bin_adam, newdata = iris_bin)

predict(fit_bin_adam, iris_bin[1:5, ], type = "class")

predict(fit_bin_adam, iris_bin[1:5, ], type = "prob")

plot(fit_bin_adam)
```

For binary classification, `type = "prob"` returns a two-column probability matrix.

## Multi-Class Classification

For multi-class classification, `metANN` uses a softmax output layer by default.

If `loss = NULL`, the default loss is categorical cross-entropy.

## Example 9: Multi-Class Classification with SBOA

```{r multiclass-classification-sboa, eval=FALSE}
fit_multi_sboa <- met_mlp(
  formula = Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = iris,
  hidden_layers = c(6),
  activation = "relu",
  optimizer = optimizer_sboa(
    pop_size = 25,
    max_iter = 40
  ),
  seed = 123,
  verbose = FALSE
)

fit_multi_sboa
summary(fit_multi_sboa)

evaluate(fit_multi_sboa, newdata = iris)

predict(fit_multi_sboa, iris[1:5, ], type = "class")

predict(fit_multi_sboa, iris[1:5, ], type = "prob")

plot(fit_multi_sboa)
```

## Example 10: Multi-Class Classification with Adam

```{r multiclass-classification-adam, eval=FALSE}
fit_multi_adam <- met_mlp(
  formula = Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = iris,
  hidden_layers = c(6),
  activation = "relu",
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 200,
    batch_size = 32
  ),
  seed = 123,
  verbose = FALSE
)

fit_multi_adam
summary(fit_multi_adam)

evaluate(fit_multi_adam, newdata = iris)

predict(fit_multi_adam, iris[1:5, ], type = "class")

predict(fit_multi_adam, iris[1:5, ], type = "prob")

plot(fit_multi_adam)
```

For multi-class classification, `type = "prob"` returns a probability matrix with one column per class.

## Prediction Types

For regression models:

```{r prediction-regression, eval=FALSE}
predict(fit_reg_adam, newdata = iris[1:5, ], type = "response")
```

For classification models:

```{r prediction-classification, eval=FALSE}
predict(fit_bin_adam, newdata = iris_bin[1:5, ], type = "class")

predict(fit_bin_adam, newdata = iris_bin[1:5, ], type = "prob")

predict(fit_bin_adam, newdata = iris_bin[1:5, ], type = "response")
```

For classification models:

- `type = "class"` returns predicted class labels,
- `type = "prob"` returns predicted class probabilities,
- `type = "response"` returns class labels by default.

## Network Visualization

The `plot_network()` function visualizes the architecture of a fitted `metANN` model or an MLP architecture object.

```{r plot-network, eval=FALSE}
plot_network(fit_multi_adam)
```

It displays input, hidden, and output layers, including the number of neurons and activation functions.

You can also plot an architecture object directly:

```{r plot-network-architecture, eval=FALSE}
arch <- mlp_architecture(
  input_dim = 4,
  layers = list(
    dense_layer(6, activation = "relu"),
    dense_layer(3, activation = "softmax")
  )
)

plot_network(arch)
```

## Model Evaluation

The `evaluate()` function computes performance metrics on new data.

For regression models:

```{r evaluate-regression, eval=FALSE}
evaluate(fit_reg_adam, newdata = iris)

evaluate(
  fit_reg_adam,
  newdata = iris,
  metrics = c("rmse", "mae", "r2")
)
```

For binary classification models:

```{r evaluate-binary, eval=FALSE}
evaluate(fit_bin_adam, newdata = iris_bin)

evaluate(
  fit_bin_adam,
  newdata = iris_bin,
  metrics = c("accuracy", "precision", "recall", "f1")
)
```

For multi-class classification models:

```{r evaluate-multiclass, eval=FALSE}
evaluate(fit_multi_adam, newdata = iris)

evaluate(
  fit_multi_adam,
  newdata = iris,
  metrics = c("accuracy", "precision", "recall", "f1")
)
```

## x-y Interface

In addition to the formula-data interface, `metANN` also supports an x-y interface.

## Example 11: Regression with x-y Interface

```{r xy-regression, eval=FALSE}
x_reg <- as.matrix(
  iris[, c("Sepal.Length", "Sepal.Width", "Petal.Length")]
)

y_reg <- iris$Petal.Width

fit_xy_reg <- met_mlp(
  x = x_reg,
  y = y_reg,
  hidden_layers = c(5),
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 100,
    batch_size = 32
  ),
  seed = 123,
  verbose = FALSE
)

fit_xy_reg
summary(fit_xy_reg)

evaluate(
  fit_xy_reg,
  newdata = x_reg,
  y_true = y_reg
)
```

## Example 12: Binary Classification with x-y Interface

```{r xy-binary-classification, eval=FALSE}
iris_bin <- iris

iris_bin$IsSetosa <- factor(
  ifelse(iris_bin$Species == "setosa", "setosa", "other")
)

x_bin <- as.matrix(
  iris_bin[, c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")]
)

y_bin <- iris_bin$IsSetosa

fit_xy_bin <- met_mlp(
  x = x_bin,
  y = y_bin,
  hidden_layers = c(5),
  optimizer = optimizer_adam(
    learning_rate = 0.01,
    epochs = 100,
    batch_size = 32
  ),
  seed = 123,
  verbose = FALSE
)

fit_xy_bin
summary(fit_xy_bin)

evaluate(
  fit_xy_bin,
  newdata = x_bin,
  y_true = y_bin
)
```

## Optimizer Information

The `optimizer_info()` function provides a short summary of an optimizer.

```{r optimizer-info, eval=FALSE}
optimizer_info("sboa")

optimizer_info(
  optimizer_adam(
    learning_rate = 0.01,
    epochs = 200,
    batch_size = 32
  )
)
```

Example output includes:

- optimizer name,
- full optimizer name,
- optimizer type,
- whether it requires gradients,
- supported interfaces,
- main parameters,
- current parameter values.

## Available Activation Functions

```{r available-activations, eval=FALSE}
available_activations()
```

Currently available activation functions are:

```r
"linear" "sigmoid" "tanh" "relu" "leaky_relu" "softmax"
```

## Available Loss Functions

```{r available-losses, eval=FALSE}
available_losses()
```

Currently available loss functions are:

```r
"mse" "mae" "huber" "log_cosh" "binary_crossentropy" "crossentropy"
```

## Available Metrics

```{r available-metrics, eval=FALSE}
available_metrics()
```

Currently available metrics are:

```r
"mse" "rmse" "mae" "r2" "accuracy" "precision" "recall" "f1"
```

## Supported Tasks

The current development version supports:

| Task | Metaheuristic optimizers | SGD | Adam |
|---|---:|---:|---:|
| General-purpose optimization without gradient | Yes | No | No |
| General-purpose optimization with gradient | No | Yes | Yes |
| MLP regression | Yes | Yes | Yes |
| Binary classification | Yes | Yes | Yes |
| Multi-class classification | Yes | Yes | Yes |

## Notes

In the current development version:

- metaheuristic optimizers can train MLPs without requiring gradients,
- SGD and Adam can optimize differentiable benchmark functions when a gradient function is supplied,
- SGD and Adam can train MLPs for regression, binary classification, and multi-class classification,
- binary classification uses sigmoid output and binary cross-entropy by default,
- multi-class classification uses softmax output and cross-entropy by default,
- classification metrics are computed using macro-averaged precision, recall, and F1.

## References

The optimization algorithms and neural network components implemented in `metANN` are based on the following key references:

- Kennedy, J., and Eberhart, R. (1995). Particle Swarm Optimization. *Proceedings of ICNN'95 - International Conference on Neural Networks*, 4, 1942--1948.
- Storn, R., and Price, K. (1997). Differential Evolution -- A Simple and Efficient Heuristic for Global Optimization over Continuous Spaces. *Journal of Global Optimization*, 11, 341--359.
- Goldberg, D. E. (1989). *Genetic Algorithms in Search, Optimization, and Machine Learning*. Addison-Wesley.
- Karaboga, D., and Basturk, B. (2007). A Powerful and Efficient Algorithm for Numerical Function Optimization: Artificial Bee Colony (ABC) Algorithm. *Journal of Global Optimization*, 39, 459--471.
- Mirjalili, S., Mirjalili, S. M., and Lewis, A. (2014). Grey Wolf Optimizer. *Advances in Engineering Software*, 69, 46--61.
- Mirjalili, S., and Lewis, A. (2016). The Whale Optimization Algorithm. *Advances in Engineering Software*, 95, 51--67.
- Rao, R. V., Savsani, V. J., and Vakharia, D. P. (2011). Teaching-Learning-Based Optimization: A Novel Method for Constrained Mechanical Design Optimization Problems. *Computer-Aided Design*, 43, 303--315.
- Fu, Y., Liu, D., Chen, J., and He, L. (2024). Secretary Bird Optimization Algorithm: A New Metaheuristic for Solving Global Optimization Problems. *Artificial Intelligence Review*, 57, 123.
- Kingma, D. P., and Ba, J. (2015). Adam: A Method for Stochastic Optimization. *International Conference on Learning Representations*.
- Nair, V., and Hinton, G. E. (2010). Rectified Linear Units Improve Restricted Boltzmann Machines. *Proceedings of the 27th International Conference on Machine Learning*, 807--814.
- Bridle, J. S. (1990). Probabilistic Interpretation of Feedforward Classification Network Outputs, with Relationships to Statistical Pattern Recognition. In *Neurocomputing: Algorithms, Architectures and Applications*, 227--236.

### Metaheuristic-Based Neural Network Training

The neural network training functionality in `metANN` is related to the literature on metaheuristic-based training of feed-forward neural networks. Representative studies include:

- Montana, D. J., and Davis, L. (1989). Training Feedforward Neural Networks Using Genetic Algorithms. *Proceedings of the 11th International Joint Conference on Artificial Intelligence*, 762--767.

- Ilonen, J., Kamarainen, J.-K., and Lampinen, J. (2003). Differential Evolution Training Algorithm for Feed-Forward Neural Networks. *Neural Processing Letters*, 17, 93--105. doi:10.1023/A:1022995128597

- Karaboga, D., and Ozturk, C. (2009). Neural Networks Training by Artificial Bee Colony Algorithm on Pattern Classification. *Neural Network World*, 19(3), 279--292.

- Mirjalili, S. (2015). How Effective is the Grey Wolf Optimizer in Training Multi-Layer Perceptrons. *Applied Intelligence*, 43, 150--161. doi:10.1007/s10489-014-0645-7

- Dilber, B., and Özdemir, A. F. (2026). A novel approach to training feed-forward multi-layer perceptrons with recently proposed secretary bird optimization algorithm. *Neural Computing and Applications*, 38(5). doi:10.1007/s00521-026-11874-x

## Citation

If you use `metANN` in academic work, please cite the package and the related optimization algorithms where appropriate.

## Authors

`metANN` is developed and maintained by:

- Burak Dilber
- A. Fırat Özdemir

## License

This package is released under the MIT license.
